Highly Available Transactions: Virtues and Limitations

### 2013-05-01 writeup/evaluation.tex
more writin

To further understand the performance implications of HAT guarantees
in a real-world environment, we implemented a HAT database
prototype. Across a range of deployments on public cloud
infrastructures, we found that, as Section~\ref{sec:latency}'s
measurements suggested, ``strongly consistent'' algorithms incur
substantial latency penalties: from 10 to 100 times their HAT
counterparts. Moreover, HAT algorithms like TA---while more
resource-intensive than basic eventual consistency---are able to scale
well to many servers.

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based, partitioned) key-value backed
by LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (last-writer-wins RU, via simple
anti-entropy) and HAT RC and TA guarantees (as described in
Section~\ref{sec:hats}). It also supports non-HAT operation whereby
all operations for a given key are routed to a (randomly) designated
\textit{master} replica for each key (i.e., PNUTS~\cite{pnuts}) as
well as distributed two-phase locking. Servers are durable: they
synchronously write to LevelDB before responding to client requests,
while pending writes in TA are sent to a write-ahead log, which is
synchronously flushed to disk.

\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
in \textit{clusters}---disjoint sets of database servers that each
contain a single, fully replicated copy of the data---typically across
datacenters, sticking all clients within a datacenter to their
respective cluster (trivially providing read-your-writes and monotonic
reads guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
instances as servers in each cluster. For our workload, we link our
client library to the YCSB benchmark~\cite{ycsb}, grouping every eight
YCSB operations from the default workload (50\% reads, 50\% writes) to
form a transaction. We increase the number of keys in the workload
from the default $1,000$ to $100,000$ with uniform random key access,
keeping the default value size of $1KB$, and running YCSB for 180
seconds per configuration.

@@

 In contrast, when the two clusters are deployed across the
 continental United States (Figure~\ref{fig:wan-exp}B), the average
 latency of mastered operation increases to $300$ms (a $278$--$4257\%$
 latency increase; average $37$ms latency per operation). For the same
 number of YCSB client threads, mastered operation has substantially
 lower throughput than the HAT configurations. Increasing the number of YCSB
 clients \textit{does} increase the throughput of mastered operation,
 but our Thrift-based connection processing did not gracefully handle
 more than several thousand concurrent connections. In contrast,
 across two datacenters, the performance of eventual, RC, and TA is
 near-identical to a single-datacenter deployment.

When we deployed five clusters (as opposed to two, as before) across
the five AWS datacenters with lowest communication cost
(Figure~\ref{fig:wan-exp}C), the trend continues: mastered latency
increases to nearly $800$ms per transaction. As an attempt at reducing
this overhead, we implemented and benchmarked a variant of
quorum-based replication as in Dynamo~\cite{dynamo}, where clients
sent requests to all replicas, which completed as soon as a majority
of servers responded (guaranteeing regular
semantics~\cite{herlihy-art}); this strategy (not pictured) did not
substantially improve performance due to the network topology and
because worst-case server load was unaffected. In this configuration,
TA's relative throughput decreased: every YCSB \texttt{put} operation
resulted in four \texttt{put} operations on remote replicas and,
accordingly, the cost of anti-entropy increased (e.g., each server
processed four incoming anti-entropy streams---as opposed to one
before---reducing the opportunity for batching and decreasing
available resources for for incoming client requests). This in turn
increased garbage collection activity and, more importantly, IOPS when
compared to Eventual and RC, causing throughput to peak at around half
of Eventual. With an in-memory persistence (i.e., no LevelDB or WAN),
TA throughput was within 20\% of eventual.

@@

\vspace{.5em}\noindent\textit{Read proportion.} The default equal
proportion of reads and writes is fairly pessimistic: for example,
Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
with all reads, TA is within $4.8\%$ of eventual; with all writes,
eventual throughput decreases by $288.8\%$ compared to all reads and
TA is within $33\%$ of eventual. At $99.8\%$ reads, TA incurs a $7\%$
overhead ($5.8\%$ for in-memory storage).

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for both eventual, RC, and TA. RC and eventual scale perfectly with an
approximately $5$x throughput increase moving from $5$ to $25$ servers
per cluster. For the same configuration, TA scales by $3.8$x,
achieving over $260,000$ operations per second. We believe that TA's
performance is due to resource contention---with a memory-backed
database (instead of LevelDB), TA scales by $4.25$x (not shown)---and
increased garbage collection activity.

>>>>

To further understand the performance implications of HAT guarantees
in a real-world environment, we implemented a HAT database
prototype. Across a range of deployments on public cloud
infrastructures, we found that, as Section~\ref{sec:latency}'s
measurements suggested, ``strongly consistent'' algorithms incur
substantial latency penalties: over WAN, 10 to 100 times higher than
their HAT counterparts. In light of recent algorithms for providing
causally consistent get transactions~\cite{bolton, cops}, we focus our
HAT analysis on TA, which we believe to be the most novel of our
algorithms. We find that HAT properties---while more
resource-intensive than basic eventual consistency---scale well to
many servers.

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based, partitioned) key-value backed
by LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (last-writer-wins RU with standard
all-to-all anti-entropy between replicas) and HAT RC and TA guarantees
(as described in Section~\ref{sec:hats}). It also supports non-HAT
operation whereby all operations for a given key are routed to a
(randomly) designated \texttt{master} replica for each key (i.e.,
PNUTS~\cite{pnuts}) as well as distributed two-phase locking. Servers
are durable: they synchronously write to LevelDB before responding to
client requests, while pending writes in TA are sent to a write-ahead
log, which is synchronously flushed to disk.

\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
in \textit{clusters}---disjoint sets of database servers that each
contain a single, fully replicated copy of the data---typically across
datacenters, sticking all clients within a datacenter to their
respective cluster (trivially providing read-your-writes and monotonic
reads guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
instances as servers in each cluster. For our workload, we link our
client library to the YCSB benchmark~\cite{ycsb}, which is well-suited
to LevelDB's key-value schema, grouping every eight YCSB operations
from the default workload (50\% reads, 50\% writes) to form a
transaction. We increase the number of keys in the workload from the
default $1,000$ to $100,000$ with uniform random key access, keeping
the default value size of $1KB$, and running YCSB for 180 seconds per
configuration.

@@

 In contrast, when the two clusters are deployed across the
 continental United States (Figure~\ref{fig:wan-exp}B), the average
 latency of mastered operation increases to $300$ms (a $278$--$4257\%$
 latency increase; average $37$ms latency per operation). For the same
 number of YCSB client threads, mastered operation has substantially
 lower throughput than the HAT configurations. Increasing the number
 of YCSB clients \textit{does} increase the throughput of mastered
 operation, but our Thrift-based server-side connection processing did
 not gracefully handle more than several thousand concurrent
 connections. In contrast, across two datacenters, the performance of
 eventual, RC, and TA is near-identical to a single-datacenter
 deployment.

When we deployed five clusters (as opposed to two, as before) across
the five AWS datacenters with lowest communication cost
(Figure~\ref{fig:wan-exp}C), the trend continues: mastered latency
increases to nearly $800$ms per transaction. As an attempt at reducing
this overhead, we implemented and benchmarked a variant of
quorum-based replication as in Dynamo~\cite{dynamo}, where clients
sent requests to all replicas, which completed as soon as a majority
of servers responded (guaranteeing regular
semantics~\cite{herlihy-art}); this strategy (not pictured) did not
substantially improve performance due to the network topology and
because worst-case server load was unaffected. With five clusters,
TA's relative throughput decreased: every YCSB \texttt{put} operation
resulted in four \texttt{put} operations on remote replicas and,
accordingly, the cost of anti-entropy increased (e.g., each server
processed four incoming anti-entropy streams---as opposed to one
before---reducing the opportunity for batching and decreasing
available resources for for incoming client requests). This in turn
increased garbage collection activity and, more importantly, IOPS when
compared to Eventual and RC, causing TA throughput to peak at around half
of Eventual. With an in-memory persistence (i.e., no LevelDB or WAL),
TA throughput was within 20\% of eventual.

@@

\vspace{.5em}\noindent\textit{Read proportion.} Our default (equal)
proportion of reads and writes is fairly pessimistic: for example,
Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
with all reads, TA is within $4.8\%$ of eventual; with all writes, TA
is within $33\%$, and Eventual throughput decreases by $288.8\%$
compared to all reads and TA is within $33\%$ of eventual. At $99.8\%$
reads, TA incurs a $7\%$ overhead ($5.8\%$ for in-memory storage).

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for both eventual, RC, and TA. RC and eventual scale perfectly with an
approximately $5$x throughput increase moving from $5$ to $25$ servers
per cluster. For the same configuration, TA scales by $3.8$x,
achieving over $260,000$ operations per second. We believe that TA's
performance is due to resource contention---with a memory-backed
database (instead of LevelDB), TA scales by $4.25$x (not shown)---and
TA-related performance heterogeneity across servers (Calvin's authors
report similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this
linear scalability is not typical of traditional database system
designs~\cite{gray-isolation}.

### 2013-05-01 writeup/hat-analysis.tex
more writin

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}, requiring only that each transaction's writes are ordered
consistently according to a single total order on transactions. This
prohibits Adya's phenomenon $G0$, also called ``Dirty
Writes''~\cite{adya}. More formally, we require a total ordering on
writes to each data item and we create a graph of transactions with
edges from one transaction to another if the former overwrites the
latter's write to the same object. \textit{PL-1} prohibits cycles in
this graph, requiring some total order on writes. Accordingly, the
final database state cannot contain the ``earlier'' transaction's
writes to the data items: for example, in the below example, $T_3$
should \emph{eventually} only read $a=b=1$ or $a=b=2$ but not $a=2,
b=1$ or $a=1, b=2$:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\vspace{-1em}
\end{align*}
Later properties will strengthen Read Uncommitted.

Unlike the total order required by serializability, the order in PL-1
does not constrain the values seen by a transaction's read
operations except once the database has reached a ``final
state.'' Read Uncommitted is easily achieved via applying the same
logical timestamp to each update in a transaction and applying a
``last writer wins'' conflict reconciliation policy at each replica.

@@

\textbf{Transactional atomicity (TA)} is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property, TA
restricts transactions' ability to view the effects of partially
completed transactions. Under TA, once some of the effects of a
transaction $T_i$ are observed by another transaction $T_j$,
thereafter all effects of $T_i$ are observed by $T_2$. Together with
item cut isolation, TA prevents Read Skew anomalies (Berenson et al.'s
A5A~\cite{ansicritique}). As an example of TA, because $T_2$ has read
$T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
for each key):
\vspace{-1em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

@@

Berenson et al. define \textit{Lost Update} as when one
transaction $T1$ reads a given data item, a second transaction $T2$
updates the same data item, then $T1$ modifies the data item based on
its original read of the data item, ``missing'' or ``losing'' $T2$'s
newer update. Consider a database containing only the following
transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: r_x(a) w_x(a+2)
\\T_2 &: w_x(2)\vspace{-1em}
\end{align*}
If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
operation, then the database will end up with $a=3$, a state that
could not have resulted in a serial execution due to $T_2$'s
``Lost Update.''

>>>>

To begin, \textbf{Read Uncommitted} isolation is captured by Adya as
\textit{PL-1}. In this model, writes to each object are totally
ordered, corresponding to the order in which they are installed in the
database. In a distributed database, different replicas may receive
writes to their local copies of data at different times, but the order
in which replicas overwrite versions should be consistent with the total order for
each item. \textit{PL-1} requires that writes to different objects are
ordered consistently across transactions, prohibiting Adya's
phenomenon $G0$, also called ``Dirty Writes''~\cite{adya}. If we
create a graph of transactions with edges from one transaction to
another if the former overwrites the latter's write to the same
object, then, under Read Uncommitted, there should be no cycles in
this graph. Consider the following example:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\end{align*}
In this example, under Read Uncommitted, it is impossible for the
database to order $T_1$'s $w_x(1)$ before $T_2$'s $w_x(2)$ but order
$T_2$'s $w_y(2)$ before $T_1$'s $w_y(1)$.  Read Uncommitted is easily
achieved via applying the same logical timestamp to each update in a
transaction and applying a ``last writer wins'' conflict
reconciliation policy at each replica. Later properties will
strengthen Read Uncommitted.

@@

\textbf{Transactional atomicity (TA)} is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property, TA
restricts transactions' ability to view the effects of partially
completed transactions. Under TA, once some of the effects of a
transaction $T_i$ are observed by another transaction $T_j$,
thereafter, all effects of $T_i$ are observed by $T_j$. Together with
item cut isolation, TA prevents Read Skew anomalies (Berenson et al.'s
A5A~\cite{ansicritique}). As an example of TA, because $T_2$ has read
$T_1$'s write to $y$, $T_2$ must observe $b=c=1$ (or later versions
for each key):
\vspace{-1em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

@@

Berenson et al. define \textit{Lost Update} as when one
transaction $T1$ reads a given data item, a second transaction $T2$
updates the same data item, then $T1$ modifies the data item based on
its original read of the data item, ``missing'' or ``losing'' $T2$'s
newer update. Consider a database containing only the following
transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: r_x(a)~w_x(a+2)
\\T_2 &: w_x(2)\vspace{-1em}
\end{align*}
If $T_1$ reads $a=1$ but $T_2$'s write to $x$ precedes $T_1$'s write
operation, then the database will end up with $a=3$, a state that
could not have resulted in a serial execution due to $T_2$'s
``Lost Update.''

### 2013-05-01 writeup/hat-appendix.tex
more writin

>>>>

\begin{algorithm}[t!]
\newcommand{\myindent}{\hspace{-1em}}
\begin{addmargin}[-1em]{0em}% 1em left, 2em right

\begin{algorithmic}
\caption*{\textbf{Algorithm} HAT Transactional Atomicity}\vspace{.5em}
\State{\textit{\textbf{Shared Data Types}}\\\vspace{.5em}
\texttt{timestamp}: unique, per transaction identifier\\
$\texttt{write}:[\texttt{key:}k, \texttt{value}:v, \texttt{timestamp}:ts, \texttt{set<key>}:tx\_keys]$\\\hrulefill}\vspace{.5em}

  \State{\textbf{\textit{Server-side Data Structures and Methods}\\\vspace{.5em}
$\texttt{set<write>:pending}$\\
$\texttt{set<write>:good}$\\
$\texttt{map<timestamp, int>:acks}$}}\\

\Procedure{put}{\texttt{write}:$w$}
  \State \texttt{pending}.add($w$)
  \For{key $k_t \in w.tx\_keys$}
  \State \{all replicas for $sib$\}.notify($w.ts$)
  \EndFor

  \State \textit{asynchronously send $w$ to other replicas via anti-entropy}
  \State \Return

  \EndProcedure\vspace{.5em}

  \Procedure{notify}{\texttt{timestamp}:$ts$}
  \State \texttt{acks}.get($ts$).increment()
  \If{all \texttt{acks} received for all replicas for $ts$'s $tx\_keys$}
  \State \texttt{good}.add($w$)
  \State \texttt{pending}.remove($w$)  
  \EndIf  
  \EndProcedure\vspace{.5em}

  \Procedure{get}{\texttt{key}:$k$, \texttt{timestamp}:$ts_{required}$}
  \If{$ts_{required} = \bot$}
  \State \Return $w \in \texttt{good}$ s.t. $w.key = key$ with highest timestamp
  \ElsIf{$\exists~w \in \texttt{good}$ s.t. $w.key = key, w.ts \geq ts_{required}$}
  \State \Return $w$
  \Else
  \State \Return $w \in \texttt{pending}$ s.t. $w.key = key,  w.ts = ts_{required}$
  \EndIf
  \EndProcedure\\\vspace{-.5em}\hrulefill\vspace{.25em}

  \State{\textbf{\textit{Client-side Data Structures and Methods}\\\vspace{.5em}
$\texttt{int:cur\_txn}$\\
$\texttt{map<key, value>:write\_buffer}$\\
$\texttt{map<key, timestamp>:required}$}}\\

\Procedure{begin\_transaction}{}
  \State \texttt{cur\_txn} = \textit{new txn ID} \commentt{// (e.g., clientID+logicalClock)}
  \EndProcedure\vspace{.5em}

\Procedure{put}{\texttt{key:$k$}, \texttt{value:$v$}}
  \State \texttt{write\_buffer}.put($k$, $v$)
\EndProcedure\vspace{.5em}

\Procedure{get}{\texttt{key:$k$}}
  \If {$k \in \texttt{write\_buffer}$}
  \State \Return \texttt{write\_buffer}.get($k$) \commentt{// if we want per-TxN RYW}
  \EndIf

  \State $w\_{ret}$ = (available replica for $k$).get($k$, \texttt{required}.get($k$))
  \For {$(tx_{key} \in w_{ret}.tx\_keys$}
  \If {$w_{ret}.ts > \texttt{required}.get(tx_{key})$}
  \State $\texttt{required}$.put$(tx_{key}, w_{ret}.ts)$
  \EndIf
  \EndFor

  \State \Return $w\_{ret}.value$
\EndProcedure\vspace{.5em}

\Procedure{commit}{}
 \For{$(tx_{key}, v) \in \texttt{write\_buffer}$}
 \State $r$ = (available replica for $tx_{key}$)
 \State $r$.put($[tx_{key}, v, \texttt{cur\_txn}, \texttt{write\_buffer}.keys()]$)
 \EndFor
  \State \textsc{cleanup()}
\EndProcedure\vspace{.5em}

\Procedure{abort}{}
  \State \textsc{cleanup()}
\EndProcedure\vspace{.5em}

\Procedure{cleanup}{}
  \State \texttt{write\_buffer}.clear()
  \State \texttt{required\_map}.clear()
\EndProcedure\vspace{.5em}

\end{algorithmic}
\end{addmargin}
\end{algorithm}
